Improving the portuguese hyphenation rules

### 2023-11-24 article/article.tex
fix Issue #31
fix Issue #32

To simplify, if we consider only the Latin alphabet, with no diacritcs, the patterns used 
in \TeX{} hyphenation are of the form: \texttt{\^{}\textbackslash{}.?[0-9]?([a-z]+[0-9]?)+\textbackslash{}.?\$},
where we have described it using a regular expression\footnote{
    Regular expressions (regex) are powerful search patterns used in text processing to find, match, and manipulate strings of text. They are a fundamental tool in programming and are supported in many programming languages.
    The regex given here uses the Perl syntax and might be broken in the following
    parts: \texttt{\^{}} and \texttt{\$} mark the start and end of the string;
    \texttt{\textbackslash{}.?} specifies an optional period; \texttt{[0-9]?} allows for an optional single digit;
    \texttt{([a-z]+[0-9]?)+} matches sequences of lowercase letters interleaved with optional digits.
}. 
One example of such pattern is \texttt{4z1z2}. Each
pattern is made of a sequence of letters and numbers. Odd numbers indicate a
good hyphenation point, whereas even numbers indicate a bad place to break.
The given example states that the sequence has a good breaking point between
the first and the second \emph{z} and an hyphenation should be inhibited before
the first \emph{z} and after the second \emph{z}. For example, the hyphenation
of the word \emph{piz-za}, \emph{fiz-zle} and \emph{mez-zanine} use this rule,
where we see the hyphen placed between the two \emph{z}'s and no hyphen before,
nor after the \emph{z}'s. Patterns may also use period symbol (\emph{.}) to
indicate word boundaries. The pattern \texttt{.sh2} applies to beginning of
words, implying that the \emph{s} and the \emph{h} should stick together in
beginning of a word and an hyphenation should also be inhibited after the
\emph{h}. For example, this pattern is used in \emph{Sher-lock}.

@@

%
% FILE="../data/default.TeX.pt-br.patterns"; grep "^1..$" "$FILE" | awk '{print substr($0, 0, 2)}' | sort | uniq -c | while read line; do p=$(echo $line | cut -d' ' -f2); c=$(grep "^$p[^0-9]$" "$FILE" | awk '{print substr($0, 3)}' | tr '\n' ' '); printf "$line $c \n"; done | cut -d' ' -f1 | paste -sd+ | bc
%
Out of the default \TeX{} rules, 252 (82\%) follow the pattern \verb|1CV|, which
represents the recurring CV syllables in Portuguese.  As for the consonants,
there are typically 18 considered (\verb|b|, \verb|c|, \verb|ç|, \verb|d|,
\verb|f|, \verb|g|, \verb|j|, \verb|k|, \verb|l|, \verb|m|, \verb|n|, \verb|p|,
\verb|r|, \verb|s|, \verb|t|, \verb|v|, \verb|x|, and \verb|z|) that can
combine with 14 vowels (\verb|a|, \verb|e|, \verb|i|, \verb|o|, \verb|u|,
\verb|á|, \verb|â|, \verb|ã|, \verb|é|, \verb|í|, \verb|ó|, \verb|ú|, \verb|ê|,
and \verb|õ|), resulting in these CV patterns that indicate favorable
hyphenation points before the consonants\footnote{The complete list of these
252 patterns is given here: \ExceptionRulesCV{}.}. 
It is worth noting that there might
be other rules or exceptions in the hyphenation patterns not covered by these
default rules.  To accommodate exceptions, 20 additional rules were created for
cases involving consonants \verb|b|, \verb|c|, \verb|d|, \verb|f|, \verb|g|,
\verb|k|, \verb|p|, \verb|t|, \verb|v|, or \verb|w| followed by \verb|l| or
\verb|r|\footnote{The 20 additional rules are: \ExceptionRulesLR{}.}
; and 3 additional rules for \verb|c|, \verb|l| or \verb|n| followed by \verb|h|\footnote{
Those 3 rules are: \ExceptionRulesCLNH{}.}.
%
% FILE="../data/default.TeX.pt-br.patterns"; grep -v "^1..$" "$FILE" | grep "2" | awk '{print substr($0, 2, 1)}' | sort | uniq | while read line; doc=$(grep "^1${line}2.$" "$FILE" | tr -d '12' | awk '{print substr($0, 2, 1)}' | tr '\n' ' '); printf "$line $c\n"; done
%
Moreover, 23 distinct patterns were introduced to indicate hyphenation points
between vowels or between \verb|c|'s, \verb|r|'s, and \verb|s|'s\footnote{
Those patterns are: \ExceptionRulesThree{}.
}. 
%
% FILE="../data/default.TeX.pt-br.patterns"; grep -v "^1..$" "$FILE" | grep -v "2" | grep "3" | wc -l
%
Additionally, 8 patterns adhere to the \verb|1[gq]u4V|, signaling beneficial
hyphenation points before \verb|g| or \verb|q|, followed by a sequence of
\verb|u| and a vowel, with an inhibiting point between the \verb|u| and the
subsequent vowel\footnote{They are the following: \ExceptionRulesFour{}.}.
%
Lastly, there is a single pattern represented as \verb|1-|, denoting that a
hyphen indeed serves as a beneficial hyphenation point."

>>>>

To simplify, if we consider only the Latin alphabet, with no diacritcs, the patterns used 
in \TeX{} hyphenation are of the form: \texttt{\^{}\textbackslash{}.?[0-9]?([a-z]+[0-9]?)+\textbackslash{}.?\$},
where we have described it using a regular expression\footnote{
    Regular expressions (regex) are powerful search patterns used in text processing to find, match, and manipulate strings of text. They are a fundamental tool in programming and are supported in many programming languages.
    The regex given here uses the Perl syntax and might be broken in the following
    parts: \texttt{\^{}} and \texttt{\$} mark the start and end of the string;
    \texttt{\textbackslash{}.?} specifies an optional period; \texttt{[0-9]?} allows for an optional single digit;
    \texttt{([a-z]+[0-9]?)+} matches sequences of lowercase letters interleaved with optional digits.
}.
One example of such pattern is \texttt{4z1z2}, which is composed of a sequence
of letters and numbers. Odd numbers indicate a
good hyphenation point, whereas even numbers indicate a bad place to break.
The given example states that the sequence has a good breaking point between
the first and the second \emph{z} and an hyphenation should be inhibited before
the first \emph{z} and after the second \emph{z}. For example, the hyphenation
of the word \emph{piz-za}, \emph{fiz-zle} and \emph{mez-zanine} use this rule,
where we see the hyphen placed between the two \emph{z}'s and no hyphen before,
nor after the \emph{z}'s. Patterns may also use period symbol (\emph{.}) to
indicate word boundaries. The pattern \texttt{.sh2} applies to beginning of
words, implying that the \emph{s} and the \emph{h} should stick together in
beginning of a word and an hyphenation should also be inhibited after the
\emph{h}. For example, this pattern is used in \emph{Sher-lock}.

@@

%
% FILE="../data/default.TeX.pt-br.patterns"; grep "^1..$" "$FILE" | awk '{print substr($0, 0, 2)}' | sort | uniq -c | while read line; do p=$(echo $line | cut -d' ' -f2); c=$(grep "^$p[^0-9]$" "$FILE" | awk '{print substr($0, 3)}' | tr '\n' ' '); printf "$line $c \n"; done | cut -d' ' -f1 | paste -sd+ | bc
%
Out of the default \TeX{} rules, 252 (82\%) follow the pattern \verb|1CV|, which
represents the recurring CV syllables in Portuguese.  As for the consonants,
there are typically 18 considered (\verb|b|, \verb|c|, \verb|ç|, \verb|d|,
\verb|f|, \verb|g|, \verb|j|, \verb|k|, \verb|l|, \verb|m|, \verb|n|, \verb|p|,
\verb|r|, \verb|s|, \verb|t|, \verb|v|, \verb|x|, and \verb|z|) that can
combine with 14 vowels (\verb|a|, \verb|e|, \verb|i|, \verb|o|, \verb|u|,
\verb|á|, \verb|â|, \verb|ã|, \verb|é|, \verb|í|, \verb|ó|, \verb|ú|, \verb|ê|,
and \verb|õ|), resulting in these CV patterns that indicate favorable
hyphenation points before the consonants\footnote{The complete list of these
252 patterns is given here: \ExceptionRulesCV{}.}. 
It is worth noting that there might
be other rules or exceptions in the hyphenation patterns not covered by these
default rules.  To accommodate exceptions, \textcite{rezende1987} proposes the 
following rules: 
\begin{itemize}
    \item 20 rules that created for cases involving consonants \verb|b|, \verb|c|, \verb|d|, \verb|f|, \verb|g|, \verb|k|, \verb|p|, \verb|t|, \verb|v|, or \verb|w| followed by \verb|l| or \verb|r|\footnote{The 20 additional rules are: \ExceptionRulesLR{}.};
    \item 3 rules for \verb|c|, \verb|l| or \verb|n| followed by \verb|h|\footnote{Those 3 rules are: \ExceptionRulesCLNH{}.}.
    % FILE="../data/default.TeX.pt-br.patterns"; grep -v "^1..$" "$FILE" | grep "2" | awk '{print substr($0, 2, 1)}' | sort | uniq | while read line; doc=$(grep "^1${line}2.$" "$FILE" | tr -d '12' | awk '{print substr($0, 2, 1)}' | tr '\n' ' '); printf "$line $c\n"; done
    \item 23 distinct patterns were introduced to indicate hyphenation points between vowels or between \verb|c|'s, \verb|r|'s, and \verb|s|'s\footnote{Those patterns are: \ExceptionRulesThree{}.}. 
    % FILE="../data/default.TeX.pt-br.patterns"; grep -v "^1..$" "$FILE" | grep -v "2" | grep "3" | wc -l
    \item 8 patterns adhere to the \verb|1[gq]u4V|, signaling beneficial hyphenation points before \verb|g| or \verb|q|, followed by a sequence of \verb|u| and a vowel, with an inhibiting point between the \verb|u| and the subsequent vowel\footnote{They are the following: \ExceptionRulesFour{}.}.
    \item 1 pattern represented as \verb|1-|, denoting that a hyphen indeed serves as a beneficial hyphenation point.
\end{itemize}
